---
title: "GVPT Maths Camp"
subtitle: "Data Visualisation"
format: 
  revealjs:
    slide-number: true
    preview-links: auto
    theme: solarized
    embed-resources: true
execute: 
  echo: true
  fig-align: "center"
---

## Learning objectives for today

1.  Introduction to R

2.  Create your first plot in R

3.  Test your hypotheses using informative data visualizations

## The Research Process

![Source: [R4DS](https://r4ds.had.co.nz/explore-intro.html)](img/r4ds_process.png){fig-align="center"}

## R basics

R code:

```{r}
#| echo: true

1 + 2
```

Functions:

```{r}
#| echo: true

sum(1, 2)
```

## EXERCISE {background-color="#B56576"}

1.  Open up RStudio.
2.  Using the console, find the summation of `45`, `978`, and `121`.
3.  What is `67` divided by `6`?
4.  What is the square root of `894`?

## CHECK YOUR ANSWERS {background-color="#B56576"}

Using the consol, find the summation of `45`, `978`, and `121`.

```{r}
sum(45, 978, 121)
```

*Or:*

```{r}
45 + 978 + 121
```

What is `67` divided by `6`?

```{r}
67 / 6
```

What is the square root of `894`?

```{r}
sqrt(894)
```

## R packages

Packages are collections of R functions, data, and compiled code in a well-defined format.

```{r}
#| echo: true
#| eval: false

# Install the relevant package(s)
install.packages("tidyverse")

# Load the packages in current session
library(tidyverse)
```

::: aside
`tidyverse` is a collection of many different packages housed within this universe. The individual packages are listed when you load the `tidyverse` package.
:::

## EXERCISE {background-color="#B56576"}

1.  Open up RStudio.
2.  Using the consol, install the `tidyverse` packages.

```{r}
#| eval: false

install.packages("tidyverse")
```

3.  Load these packages in your current session

```{r}
#| eval: false

library(tidyverse)
```

## RStudio Projects

*For your sanity's sake, for your co-author's sanity's sake*

<br>

Keeps everything:

-   Organised

-   Reproducible

-   Sustainable

::: aside
For later reference, head to Hadley Wickham's [*Workflow: projects* chapter](https://r4ds.had.co.nz/workflow-projects.html#workflow-projects) in R4DS.
:::

## Data visualisation

From [R4DS - Data Visualization](https://r4ds.had.co.nz/data-visualisation.html):

> Do cars with big engines use more fuel than cars with small engines?

::: notes
Let's use our first graph to answer a question: Do cars with big engines use more fuel than cars with small engines? You probably already have an answer, but try to make your answer precise. What does the relationship between engine size and fuel efficiency look like? Is it positive? Negative? Linear? Nonlinear?
:::

## Skipping to the end

```{r}
#| echo: false

library(ggplot2)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) + 
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  ) + 
  labs(
    title = "Relationship between engine displacement and highway miles per gallon by class",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon",
    color = "Class"
  )
```

## How did we do this?

```{r}
#| eval: false

library(ggplot2)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) + 
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  ) + 
  labs(
    title = "Relationship between engine displacement and highway miles per gallon by class",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon",
    color = "Class"
  )
```

::: aside
We will step through this now. By the end of this session, you will be able to make your own colourful data visualisations using R.
:::

## Load relevant packages and data

```{r}
#| eval: false

# Load the relevant packages
library(tidyverse)

# Load the data
mpg
```

```{r}
#| eval: true
#| echo: false

ggplot2::mpg |> 
  head() |> 
  dplyr::select(1:5) |> 
  knitr::kable()
```

::: notes
You only need to install a package once, but you need to reload it every time you start a new session.

A data frame is a rectangular collection of variables (in the columns) and observations (in the rows). mpg contains observations collected by the US Environmental Protection Agency on 38 models of car.
:::

## EXERCISE {background-color="#B56576"}

<br>

Learn more about this data set by typing `?mpg` into your console.

::: aside
To get help on any function or data set use: `?function`
:::

## Plot your data

```{r}
#| echo: true

library(ggplot2)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

::: notes
The plot shows a negative relationship between engine size (`displ`) and fuel efficiency (`hwy`). In other words, cars with big engines use more fuel. Does this confirm or refute your hypothesis about fuel efficiency and engine size?

With `ggplot2`, you begin a plot with the function `ggplot()`. `ggplot()` creates a coordinate system that you can add layers to. The first argument of `ggplot()` is the dataset to use in the graph. So `ggplot(data = mpg)` creates an empty graph, but it's not very interesting so I'm not going to show it here.

You complete your graph by adding one or more layers to `ggplot()`. The function `geom_point()` adds a layer of points to your plot, which creates a scatterplot. `ggplot2` comes with many geom functions that each add a different type of layer to a plot. You'll learn a whole bunch of them throughout this chapter.

Each geom function in `ggplot2` takes a mapping argument. This defines how variables in your dataset are mapped to visual properties. The mapping argument is always paired with `aes()`, and the `x` and `y` arguments of `aes()` specify which variables to map to the `x` and `y` axes. `ggplot2` looks for the mapped variables in the data argument, in this case, `mpg`.
:::

## EXERCISE {background-color="#B56576"}

What do you see when you run the following?

```{r}
#| echo: true
#| eval: false

ggplot(data = mpg)
```

How many rows are in `mpg`? How many columns?

```{r}
#| echo: true
#| eval: false

nrow(mpg)
ncol(mpg)
```

What does the `drv` variable describe?

```{r}
#| echo: true
#| eval: false

?mpg
```

## EXERCISE {background-color="#B56576"}

Make a scatterplot of `hwy` vs `cyl`.

```{r}
#| eval: false

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = hwy, y = cyl))
```

What happens if you make a scatterplot of `class` vs `drv`? Why is the plot not useful?

```{r}
#| eval: false

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = class, y = drv))
```

## Let's look at groups in the data

-   We are not restricted to looking at only two interesting elements of our data.

-   You can use visual elements or aesthetics (`aes`) to communicate many dimensions in your data.

-   Let's look at a categorical variable: the class of car (SUV, 2 seater, pick up truck, etc.).

-   *Look for meaningfully defined groups.*

## Let's look at groups in the data

```{r}
#| code-line-numbers: "2"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

## Let's look at groups in the data

```{r}
#| code-line-numbers: "2"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

## Let's look at groups in the data

```{r}
#| code-line-numbers: "2"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

## Flexible visualization

You can use visual elements to communicate your findings in engaging ways.

```{r}
#| code-line-numbers: "2"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class == "2seater"))
```

## Changing the look of your plots

```{r}
#| code-line-numbers: "2"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "red")
```

## EXERCISE {background-color="#B56576"}

What's gone wrong with this code? Why are the points not blue?

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = "blue"))
```

## EXERCISE {background-color="#B56576"}

Which variables in `mpg` are categorical? Which variables are continuous?

<br>

Map a continuous variable to `color`, `size`, and `shape.` How do these aesthetics behave differently for categorical vs. continuous variables?

<br>

What happens if you map the same variable to multiple aesthetics?

## Let's add useful headings

```{r}
#| code-line-numbers: "3-7"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "red") + 
  labs(
    title = "Relationship between engine displacement and highway miles per gallon",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

## Let's clean this up

Less is more when it comes to data visualization.

```{r}
#| eval: false
#| code-line-numbers: "3"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") + 
  theme_minimal() + 
  labs(
    title = "Relationship between engine displacement and highway miles per gallon",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

::: aside
All pre-set `ggplot` themes can be found in the [`ggplot` documentation](https://ggplot2.tidyverse.org/reference/ggtheme.html).
:::

## Let's clean this up

```{r}
#| echo: false

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy), color = "blue") + 
  theme_minimal() + 
  labs(
    title = "Relationship between engine displacement and highway miles per gallon",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon"
  )
```

## Creating your own theme

```{r}
#| eval: false
#| code-line-numbers: "3-9"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) + 
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  ) + 
  labs(
    title = "Relationship between engine displacement and highway miles per gallon by class",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon",
    color = "Class"
  )
```

::: aside
All theme elements that you can control can be found in the [`ggplot` documentation](https://ggplot2.tidyverse.org/reference/theme.html#arguments). There are a lot!
:::

## Creating your own theme

```{r}
#| echo: false

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class)) + 
  theme(
    legend.position = "bottom",
    panel.grid = element_blank(),
    panel.background = element_blank(),
    plot.title.position = "plot",
    plot.title = element_text(face = "bold")
  ) + 
  labs(
    title = "Relationship between engine displacement and highway miles per gallon by class",
    x = "Engine displacement (L)",
    y = "Highway miles per gallon",
    color = "Class"
  )
```

## EXERCISE {background-color="#B56576"}

1.  Create a scatterplot of `hwy` vs `displ` and a categorical variable in the `mpg` data set.

2.  Customize your plot using the `theme()` argument.

## Looking at group-specific trends

```{r}
#| code-line-numbers: "3"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

## Looking at group-specific trends

```{r}
#| code-line-numbers: "3"

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

## EXERCISE {background-color="#B56576"}

1.  What happens if you facet on a continuous variable?

2.  What do the empty cells in previous plot with facet_grid(drv \~ cyl) mean? How do they relate to this plot?

```{r}
#| echo: false

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = drv, y = cyl))
```

## EXERCISE {background-color="#B56576"}

3.  Create the following plots. What does `.` do?

```{r}
#| eval: false

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(drv ~ .)

ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  facet_grid(. ~ cyl)
```

## Summarising relationships in your data

```{r}
#| echo: true

ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = F)
```

::: notes
Because both `geom_point()` and `geom_smooth()` take the argument `mapping = aes(x = displ, y = hwy)`, we can feed it to them both through `ggplot()`. We are previewing a feature heavily used in the tidyverse: the pipe.
:::

## Group-specific relationships

```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy, color = drv)) + 
  geom_point() + 
  geom_smooth(method = "lm", se = F) + 
  theme_minimal()
```

## Summary

This session you:

1.  Set up your data science tools

2.  Plotted complex data in an engaging way

3.  Discovered interesting relationships in the data

4.  Connected these relationships or trends to your expectations (or hypotheses about the data)
